---
title: "Assignment 01"
output: html_document
Student: "Jialiang Wu"
unique name: "alfredwu"
---

```{r setup, include = TRUE}
indata <- read.csv("D:/RStudio/File/STATS413/vwjetta_gas.csv")
indata$mileage[indata$mileage == 999999] <- NA
indata$vehage[indata$vehage == 999] <- NA
indata$mpg[indata$mpg == 99] <- NA

indata$congrade <- factor(indata$congrade)
levels(indata$congrade) <- c("rough", "average", "clean","excellent")
```

## Question a

* Use the summary function to get variable summaries after recoding missing values and declaring congrade as a factor with four levels
* How many values are missing on mileage? on vehage? on mpg?

```{r question a, include = TRUE}
summary(indata)
```
* 5 values are missing in **mileage**
* 8 values are missing in **vehage**
* 6 values are missing in **mpg**

##  Question b

* How many vehicles are there with condition grade equal to rough, average, clean, and excellent? In other words, give the frequency distribution of congrade.

```{r question b, include = TRUE}
table(indata$congrade)
barplot(table(indata$congrade))
```

## Question c

* Make a matrix scatter plot of the variables resratio, mileage, and vehage.

```{r, include = TRUE}
pairs(indata[,c("resratio", "mileage", "vehage")])
```

## Question d

* Make side-by-side boxplots of resratio by the levels of congrade. Calculate the mean resratio by the levels of congrade.

```{r, include = TRUE}
resratio_rough <- indata$congrade == "rough"
resratio_average <- indata$congrade == "average"
resratio_clean <- indata$congrade == "clean"
resratio_excellent <- indata$congrade == "excellent"

boxplot(indata$resratio[resratio_rough], indata$resratio[resratio_average], indata$resratio[resratio_clean], indata$resratio[resratio_excellent])
```

